---
format:
  revealjs:
    incremental: true
    slide-number: true
    progress: true
    code-line-numbers: true
    code-copy: true
    scrollable: true
    theme:
      - EDS213_slides.scss
    embed-resources: true
---

##  {#title-slide data-menu-title="Title Slide" background="#cad2c5"}

[EDS 213 Lab 2]{.custom-title}

<hr class="hr-off-white">

[Data Cleaning]{.custom-subtitle}


##  {#overview data-menu-title="Overview"}
[Today's Lab]{.slide-title}




Real-world datasets are messy. We'll walk through **three example databases** and the cleaning steps each one needs before ingesting into a database.





```{r}
#| echo: false
library(tidyverse)
library(dplyr)
library(lubridate)
library(kableExtra)

# --- Database 1 ---
air_quality <- data.frame(
  Date     = c("2020-01-01", "2020/02/01", "March 3, 2020", NA, "2020-04-01"),
  Location = c("Site A", "Site B", "Site C", "Site A", "Site D"),
  PM2_5    = c(12.3, "fifteen", 7.8, 9.0, 10.2),
  NO2      = c(20, 30, 25, "NA", 40),
  O3       = c(15, 20, "twenty", 30, 35)
)

water_quality <- data.frame(
  Sample_Date        = c("01-01-2020", "2020-02-01", "03-03-2020", "2020-Apr-04", "2020-05-01"),
  Site_Name          = c("River X", "Lake Y", "River X", "Lake Z", "River X"),
  pH                 = c(6.5, 7.0, 7.5, "7.2", 8.1),
  Turbidity          = c(3.2, 2.8, NA, 3.5, 300),
  Lead_Concentration = c(0.05, 0.07, "0.1", 0.04, "0.030 ppm")
)

# --- Database 2 ---
air_quality2 <- data.frame(
  Date     = c("2023-04-01", "2023-04-01", "2023-04-02", "2023-04-02", "2023-04-03"),
  Time     = c("08:00", "14:00", "08:00", "14:00", "08:00"),
  Location = c("City Park", "City Park", "Industrial Area", "Industrial Area", "Residential Zone"),
  PM10     = c(50, 55, 150, "155 mg/m³", 40),
  PM2_5    = c("25 µg/m³", 30, 75, 80, NA),
  NO2      = c(40, 42, 60, NA, 35)
)

daily_weather <- data.frame(
  Location = c("Park", "Beach", "Forest", "Hill"),
  `2023-04-01 Temp 06:00`     = c(14, 16, 13, 15),
  `2023-04-01 Humidity 06:00` = c(80, 85, 90, 75),
  `2023-04-01 Temp 12:00`     = c(22, 24, 21, 23),
  `2023-04-01 Humidity 12:00` = c(60, 55, 65, 50),
  `2023-04-01 Temp 18:00`     = c(20, 22, 19, 21),
  `2023-04-01 Humidity 18:00` = c(70, 75, 80, 65),
  `2023-04-01 Temp 24:00`     = c(16, 18, 15, 17),
  `2023-04-01 Humidity 24:00` = c(85, 90, 95, 80)
)

# --- Database 3 ---
set.seed(123)
dates     <- seq(as.Date('2020-01-01'), as.Date('2022-12-31'), by = 'day')
locations <- c("Location1", "Location2", "Location3")
n         <- length(dates) * length(locations)

daily_air_quality <- data.frame(
  Date     = rep(dates, each = length(locations)),
  Location = rep(locations, times = length(dates)),
  PM2_5    = runif(n, 5, 150),
  NO2      = runif(n, 10, 100),
  Ozone    = runif(n, 20, 120)
)

years <- 2020:2022
yearly_vegetation_health <- expand.grid(Year = years, Location = locations)
yearly_vegetation_health$VHI <- runif(nrow(yearly_vegetation_health), 40, 100)
```


##  {#Database1 data-menu-title="Database # 1"}
[Database #1]{.slide-title}

:::: {.columns}

::: {.column width="55%"}
```{r}
#| echo: false
knitr::kable(air_quality, caption = "Air Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```
:::

::: {.column width="45%"}
```{r}
#| echo: false
knitr::kable(water_quality, caption = "Water Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```
:::

::::


##  {#Database1-issues data-menu-title="Database # 1 Issues"}
[Database #1: What's Wrong?]{.slide-title}

**Air Quality table:**

- `Date` — mixed formats: `2020-01-01`, `2020/02/01`, `March 3, 2020`
- `PM2_5` — contains `"fifteen"` instead of a number
- `NO2` — contains `"NA"` as a string, not a true missing value

**Water Quality table:**

- `Sample_Date` — multiple inconsistent date formats
- `Lead_Concentration` — mixes plain numbers with `"0.030 ppm"`




##  {#Database2 data-menu-title="Database # 2"}
[Database #2]{.slide-title}

:::: {.columns}

::: {.column width="45%"}
```{r}
#| echo: false
knitr::kable(air_quality2, caption = "Air Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```
:::

::: {.column width="55%"}
```{r}
#| echo: false
knitr::kable(daily_weather, caption = "Weather (wide format)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11)
```
:::

::::


##  {#Database2-issues data-menu-title="Database # 2 Issues"}
[Database #2: What's Wrong?]{.slide-title}

**Air Quality table:**

- `PM10` has one entry with `"155 mg/m³"` — unit embedded in value
- `PM2_5` has one entry with `"25 µg/m³"` — same problem

**Weather table:**

- Data is in **wide format** — each column encodes date, variable, and time
- Relational databases need **long/tidy format**
- Need `pivot_longer()` to reshape into separate `Date`, `Time`, `Temp`, and `Humidity` columns




##  {#Database3 data-menu-title="Database # 3"}
[Database #3]{.slide-title}

:::: {.columns}

::: {.column width="60%"}
```{r}
#| echo: false
knitr::kable(head(daily_air_quality), caption = "Daily Air Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```
:::

::: {.column width="40%"}
```{r}
#| echo: false
knitr::kable(yearly_vegetation_health, caption = "Yearly Vegetation Health") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```
:::

::::


##  {#Database3-issues data-menu-title="Database # 3 Issues"}
[Database #3: What's Wrong?]{.slide-title}

**Daily Air Quality table:**

- Recorded at **daily** resolution

**Yearly Vegetation Health table:**

- Recorded at **yearly** resolution
- These tables cannot be joined directly — they have different temporal resolutions

**Fix:** Aggregate daily air quality to yearly averages so both tables share a `Year` + `Location` key.


##  {#Database3-cleaning data-menu-title="Database # 3 Cleaning"}
[Database #3 Cleaning]{.slide-title}

```{r}
yearly_air_quality <- daily_air_quality %>%
  group_by(Year = lubridate::year(Date), Location) %>%
  summarise(
    Avg_PM2_5 = mean(PM2_5, na.rm = TRUE),
    Avg_NO2   = mean(NO2,   na.rm = TRUE),
    Avg_Ozone = mean(Ozone, na.rm = TRUE),
    .groups = 'drop'
  )

knitr::kable(head(yearly_air_quality), caption = "Yearly Air Quality") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```


##  {#Database3-join data-menu-title="Database # 3 Join"}
[Database #3: Joining the Tables]{.slide-title}

```{r}
combined <- yearly_air_quality %>%
  left_join(yearly_vegetation_health, by = c("Year", "Location"))

knitr::kable(head(combined), caption = "Joined: Air Quality + Vegetation Health") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 14)
```

Both tables now share the same `Year` + `Location` key — the join works cleanly.


##  {#this-weeks-task data-menu-title="This Week's Task"}
[This Week's Task]{.slide-title}

::: {style="font-size: 0.8em;"}
Apply these principles to **your own dataset**. Before next week's ingestion step, make sure your data meets the following criteria:

- All date columns use a **consistent format**
- Numeric columns contain **only numbers** (no units or text mixed in)
- Missing values are represented consistently as **`NA`**
- Column names are **consistent across tables** that will be joined
- Data is in **long/tidy format** (each variable in its own column, each observation in its own row)
- Tables that will be joined share a **common key**
- Any **other cleaning steps** specific to your dataset (e.g., removing duplicates, handling outliers, standardizing category names)
:::
