---
title: "K-Anonymity with sdcMicro: Whale Entanglement Data"
format: html
editor: visual
execute:
  eval: false
---

# Background

This dataset (`whale-entanglement.csv`) contains documented whale entanglement incidents on the U.S. West Coast. Each row is one event, with variables for whale species, location, gear type, and fishery involved. *(Adapted for instructional use only.)*

**Goals:**

1.  Assess re-identification risk in the raw data
2.  Apply Statistical Disclosure Control (SDC) to reduce that risk
3.  Quantify the resulting information loss

------------------------------------------------------------------------

# Step 1: Load Package & Data

```{r}
#| label: load

```

## Inspect Data

```{r}
#| label: explore

```

Each row is one entanglement event.

------------------------------------------------------------------------

# Step 2: Identify Disclosure Risks

## Direct Identifiers

> **Direct identifiers** uniquely identify a record on their own.

**Q1 — How many direct identifiers are present, and what are they?**

<!-- TODO: Fill in your answer below. Look at the column names — which variables would uniquely identify a specific event or operator on their own? -->

**Answer:** *Your answer here.*

| Variable | Why it directly identifies |
|----------|----------------------------|
| `___`    | \_\_\_                     |
| `___`    | \_\_\_                     |
| `___`    | \_\_\_                     |

These must be **excluded** from any analytical release.

## Quasi-Identifiers

> **Quasi-identifiers (QIDs)** don't identify records alone, but can be *combined* with each other — or with external data — to re-identify individuals.

**Q2 — Which variables are quasi-identifiers?**

<!-- TODO: Which columns, when combined, could allow an adversary to re-identify a vessel or operator? Think about what's publicly searchable (e.g., state fisheries databases). -->

**Answer:** *Your answer here. List the variables and explain why each is a QID.*

------------------------------------------------------------------------

# Step 3: Prepare Variable Types

> **Q3 — What types do the variables need to be?**

`sdcMicro` requires categorical QIDs to be `factor` type and continuous variables to be `numeric`. Misspecified types produce incorrect risk estimates.

> 💡 **Tip — `varToFactor()` and `varToNumeric()`:** These `sdcMicro` helper functions convert columns in a data frame to `factor` or `numeric`. Both take `obj` (the data frame) and `var` (a character vector of column names). Work on a **copy** of your data frame so the original stays intact for later comparison.

```{r}
#| label: recode-types

# Work on a copy so the raw 'whale' object stays intact for comparison
file <- whale

# TODO: Convert categorical variables to factor using varToFactor()
# Hint: include type, county, state, condition, origin, gear, fine, infraction_type
file <- varToFactor(
  obj = file,
  var = c(___)   # <-- fill in the variable names
)

# TODO: Convert continuous variables to numeric using varToNumeric()
# Hint: year and month should be numeric
file <- varToNumeric(
  obj = file,
  var = c(___)   # <-- fill in the variable names
)
```

------------------------------------------------------------------------

# Step 4: Measure Initial Risk

## 4.1 Create the SDC Object

> **Q4.1 — What is the re-identification risk for this dataset?**

We create an `sdcMicroObj` that encodes our design decisions: which variables are QIDs, which direct identifiers to exclude, and any weights or strata.

> 💡 **Tip — `createSdcObj()`:** The key arguments are:
>
> -   `dat` — your data frame (use the **original** `whale`, not `file`)
> -   `keyVars` — character vector of quasi-identifier column names
> -   `excludeVars` — character vector of direct identifier column names to drop
>
> All other arguments (`weightVar`, `hhId`, etc.) can be `NULL` for this exercise.

```{r}
#| label: create-sdc-object

# TODO: Create the sdcMicro object. Choose your QIDs and direct identifiers.
sdcInitial <- createSdcObj(
  dat         = ___,                        # <-- which data frame?
  keyVars     = c(___),                     # <-- your QIDs
  excludeVars = c(___),                     # <-- direct identifiers to exclude
  weightVar   = NULL,
  hhId        = NULL,
  strataVar   = NULL,
  pramVars    = NULL,
  seed        = 0,
  randomizeRecords = FALSE,
  alpha       = c(1)
)
```

Then, compute the current disclosure risk:

> 💡 **Tip:** The global re-identification risk is stored in `sdcInitial@risk$global$risk`. Values closer to 1 mean higher risk.

```{r}
#| label: initial-risk

# TODO: Print the global re-identification risk
print(___)
```

## 4.2 Assess k-Anonymity Violations

> **Q4.2 — To what extent does this dataset violate k-anonymity?**

A dataset satisfies **k-anonymity** if every QID combination appears in *at least k* records. Records in smaller groups are "at risk."

> 💡 **Tip — `print()` on an `sdcMicroObj`:** Calling `print(sdcInitial)` produces a full risk report, including the number of records violating 2-, 3-, and 5-anonymity. Look for the **"Frequency of key"** and **"Number of observations violating k-anonymity"** sections.

```{r}
#| label: k-anon-violations

# TODO: Print the full risk report for sdcInitial
___
```

**Our target:** achieve at least **3-anonymity** (k = 3).

**Q4.2 answer — What share of records currently violate 3-anonymity?** *Your answer here.*

------------------------------------------------------------------------

# Step 5: Apply Anonymization

## 5.1 Non-Perturbative Method Testing — Recode `origin`

> **Q5.1 — Apply one non-perturbative method. How effective was it?**
>
> Reminder:
>
> -   **Non-perturbative methods** generalize or suppress values without adding noise.
> -   **Perturbative methods** (e.g., PRAM, microaggregation) alter values by adding controlled noise.

We can **recode** `origin` by merging minority categories into a single broader category. This reduces the distinctiveness of those records.

> 💡 **Tip — `groupAndRename()`:** Collapses one or more levels of a factor QID into a new label. Arguments:
>
> -   `obj` — your `sdcMicroObj`
> -   `var` — the variable name (as a string)
> -   `before` — character vector of original levels to merge
> -   `after` — character vector with the new label (single value repeated for all merged levels)
>
> Always check `table(sdcInitial@manipKeyVars$origin)` before and after to confirm the recode worked.

```{r}
#| label: recode-origin

# Before: check current distribution
cat("Distribution of 'origin' before recoding:\n")
print(table(sdcInitial@manipKeyVars$origin))

# TODO: Recode origin — merge "recreational" and "tribal" into "non-commercial"
sdcInitial <- groupAndRename(
  obj    = sdcInitial,
  var    = "___",          # <-- variable name
  before = c(___, ___),    # <-- levels to merge
  after  = c("___")        # <-- new label
)

# After: check new distribution
cat("\nDistribution of 'origin' after recoding:\n")
print(table(sdcInitial@manipKeyVars$origin))

# Recompute risk
cat("\nRisk summary after recoding 'origin':\n")
print(sdcInitial)
```

**Q5.1 answer — Did the recoding reduce k-anonymity violations? Why or why not?**

<!-- TODO: Reflect on whether this method was effective. What is the bottleneck? -->

*Your answer here.*

## 5.2 Automated k-Anonymization — `kAnon()`

> **Q5.2 — Apply k = 3 anonymization.**

`kAnon()` uses **local suppression**: it sets the minimum number of QID values to `NA` until every combination appears at least *k* times, minimizing total information loss.

> 💡 **Tip — `kAnon()`:** Takes your `sdcMicroObj` and `k` (an integer vector, e.g., `c(3)`). It returns a new `sdcMicroObj` with suppressed values applied. After running it, call `print()` again to see the updated violation counts — they should drop to zero for your chosen k.

```{r}
#| label: kanon

# TODO: Apply k = 3 anonymization and store the result in sdcAnon
sdcAnon <- kAnon(___, k = c(___))

cat("Risk summary after k = 3 anonymization:\n")
print(sdcAnon)
```

**Q5.2 answer — How many records still violate 3-anonymity after running `kAnon()`?** *Your answer here.*

------------------------------------------------------------------------

# Step 6: Measure Information Loss

> **Q6 — How much information did we lose?**

Anonymization is a trade-off: privacy protection at the cost of analytical utility. Quantifying this trade-off is essential for responsible data sharing.

> 💡 **Tip — `print(sdcAnon, "ls")`:** Passing `"ls"` as a second argument to `print()` displays the **local suppression summary**: how many values were suppressed in each QID column and as an overall percentage. Higher suppression = more information loss.

```{r}
#| label: info-loss

# TODO: Print the information-loss summary
print(___, "ls")
```

**Q6 answer — Which variable had the most suppression, and what does that tell you about the data?** *Your answer here.*

------------------------------------------------------------------------

# Step 7: Export the Anonymized Dataset

Before sharing, we must:

1.  **Remove direct identifiers** (already excluded from the SDC object)
2.  **Generate a new random ID** to replace `case_id` — never reuse the original, as it may link to external records

```{r}
#| label: export

# Anonymized quasi-identifier columns
anon_key_vars <- sdcAnon@manipKeyVars

# Non-key, non-excluded variables (passed through unchanged)
safe_vars <- whale[, !(names(whale) %in%
                         c("case_id", "lat", "long", "fishery_license",
                           "state", "origin", "county"))]

# Combine
anon_data <- cbind(anon_key_vars, safe_vars)

# Generate synthetic random IDs
set.seed(42)
anon_data$anon_id <- paste0("ID_", sample(100000:999999, nrow(anon_data), replace = FALSE))

# Put the new ID first
anon_data <- anon_data[, c("anon_id", setdiff(names(anon_data), "anon_id"))]

# Preview and save
head(anon_data)
write.csv(anon_data, "whale-entanglement-anon.csv", row.names = FALSE)
```

------------------------------------------------------------------------

**Key takeaway:** k-anonymity via local suppression is a principled, auditable approach to protecting microdata. The `sdcMicro` package makes every suppression traceable, and the information-loss metrics help justify trade-offs to data stakeholders.
